Liveness-Based Garbage Collection for Lazy Languages

### 2015-03-22 paper/paper.tex
Added unit tests and fixed run_in_loop.py

program shown, there are three such paths $EP_1 = \pi_1, \pi_2, \pi_3,
\pi_4,  \pi_5$, $EP_2  =  \pi_1, \pi_2,  \pi_3,  \pi_4, \pi_6,  \pi_7,
\pi_8$, and $EP_3 = \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.

The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
path to a common liveness environment, and a second map $\cal{P}$ maps
a program  point to the set of  e-paths on which  the point lies.
If a  program  point happens to fall  on more than one  such path, then
the  liveness environment of  the program  point is  the variable-wise
union of  the liveness environments  of the individual paths.   In the
example program, ${\cal M}(EP_1)= \ldots$, ${\cal M}(EP_2)=\ldots$ and
${\cal M}(EP_3)=  \ldots$.  Since  ${\cal P}(pi_4)$ is  $\lbrace EP_1,
EP_2, EP_3 \rbrace$, the liveness  of $pi_4$ is \ldots.  In summary,
our propagation of liveness is flow-insensitive but path-sensitive.
In the sequel
we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
$\cal{P}$ is easy and we do not elaborate it any further.

@@

\section{Generating Liveness Equations}
\begin{figure*}[t!]
\begin{picture}(100,130)(0,-90)
  \begin{tabular}{cc}
    \begin{boxedminipage}{.5\textwidth}
      {\sf
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\pi_3\!\!:\,$ \px
            \UNL{4} \hspace*{1.15cm} $\pi_4\!\!:\, (\RETURN_1~0)$
            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\, (\RETURN_2~(1~+~\py))$)))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $\pi_\mainpgm\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
	  \UNL{2}   \hspace*{.78cm}              (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{3}   \hspace*{.78cm}    $\pi_8\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\length\ \py) \IN~$\pi_{9}\!\!:\,$ (\RETURN\ \pw)))))))
\end{uprogram}}
        } 
    \end{boxedminipage}
    & 
    \begin{boxedminipage}{.5\textwidth}
      {\sf
        \begin{tabular}{lc}
        $\cal{M} = $ 
            \begin{tabular}{lc}            
              $[$
              $\pi_{\mbox{7}} \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
              \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
              $\pi_{\mbox{4}} \mapsto \lbrace \rbrace , $\\
              $\pi_{\mbox{3}} \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
              $]$
            \end{tabular} \\
            $\cal{P} = [$
              \begin{tabular}{lc}
              $\pi_{\mbox{1}} \mapsto \lbrace \pi_{\mbox{3}},\pi_{\mbox{4}}, \pi_{\mbox{7}} \rbrace, $ \
              $ \pi_{\mbox{2}} \mapsto \lbrace \pi_{\mbox{3}},\pi_{\mbox{4}}, \pi_{\mbox{7}} \rbrace, $ \\
              $ \pi_{\mbox{3}} \mapsto \lbrace \pi_{\mbox{3}} \rbrace, $ \
              $ \pi_{\mbox{4}} \mapsto \lbrace \pi_{\mbox{4}} \rbrace, $ \
              $ \pi_{\mbox{5}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace, $ \\
              $ \pi_{\mbox{6}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace, $ \
              $ \pi_{\mbox{7}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace $ \\
              $]$
            \end{tabular}
        \end{tabular}
      }
    \end{boxedminipage}
    \\
    (a) Example program.&
    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
        (b) {\em e-path} to liveness environment map \\
        (c) Prog pt to {\em e-paths } map
    \end{tabular}}
  \end{tabular}
\end{picture}
\kern -3ex

\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
\end{figure*}

@@

\subsection{Generating Liveness Equations for Function bodies}
For the running example, $\length$ has calls from $\mainpgm$ at $\pi_8$
and a recursive call at $\pi_6$.
\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
So $\sigma_{\length} =
     \deltacall{\length}{\pi_8}{\mainpgm}  \cup \deltacall{\length}{\pi_6}{\length}$.
Filling in  the values gives:

@@

For the running example, containing unique GC points (corresponding to unique {\em e-paths} sets) $\pi_3, \pi_4, \pi_7$ in $\length$ and
$\pi_9$ in $e_\mainpgm$, this gives

\begin{align*}
\Lanv{7}{\xl} &= \acdr\Lf{\length}{1}{\epsilon} \cup \epsilonset \\
\Lanv{7}{\pu} &= \Lf{\length}{1}{\epsilon} \\
\Lanv{7}{\py} &= \epsilonset \\
\Lanv{3}{\xl} &= \epsilonset \\
\Lanv{3}{\px} &= \epsilonset 
\\[1ex]
  \Lanv{9}{\pb}  &= \Lf{\length}{1}{\sigma_{\mathit{\!all}}} \\
  \Lanv{9}{\pw}  &= \sigma_{\mathit{\!all}}
\end{align*}

@@

\section{Implementation}
% GC assumptions
By the  time the program  starts executing, the liveness  automata are
created and stored in a file.   This is re-used every time the program
executes.
% GC points - mention reduction of gc points due to e-paths
An important  difference in GC of  lazy language compared  to an eager
language  is  the  point  of  garbage collection.  In  case  of  eager
languages only \CONS~ points and  function calls are the GC points. In
a lazy language  every \LET~ statement is a  potential GC point. Hence
we  need  to store  liveness  automata  corresponding  to every  \LET~
statement in the program. This leads  to an explosion in the number of
automata. From our observation that  along an {\em E-path} all program
points have  the same liveness, we  can reduce the  number of automata
state required  by sharing  the same automata  for all  program points
along an {\em E-path}. In the  example program, we only need to create
automata for program  points \ldots, since they are  the only distinct
{\em E-path}s in the program.
% Abstract description of the GC algorithm
\begin{algorithmic}
\STATE $new\_cell\gets$ dup\_heapcell(cell)
\IF{$celltype\ is\ cons\ and\ not(copied\_using\_rgc)$}
\STATE $new\_car\gets$ copy($cell.car$)
\STATE $new\_cell.car\gets$ $new\_car$ 
\STATE $new\_cdr\gets$ copy(cell.cdr)
\STATE $new\_cell.cdr\gets$ $new\_cdr$
\ELSE
\IF {$celltype\ is\ closure\ and\ cell\ not\ already\ copied$}
\STATE $new\_cell\gets$ dup\_heapcell(cell)
\STATE $new\_cell.arg1\gets$ copy($cell.arg1$)  
\STATE $new\_cell.arg2\gets$ copy($cell.arg2$)  
\ELSE
\STATE $new\_cell\gets$ dup\_heapcell(cell)
\ENDIF
\ENDIF
\end{algorithmic}

@@

Another  difference that is  present in  GC of  lazy language  is the
presence  of  a print  stack.  In  lazy  languages the  evaluation  of
closures is  driven by the print  function.  Hence, during  a GC there
will be heap  references that will be present on  the stack which also
have  to  copied/collected.   One  solution  is  to  assume  that  all
references  present  on  the  print  stack are  fully  live  and  copy
everything pointed  by them  akin to reachability.   This is  safe but
copies heap  cells which  have already been  printed. Since  the print
function is  not part of  the executing program its  liveness automata
are not  considered. Therefore we  simulate the liveness of  the print
function by using a flag in  the heap cell. The flag ensures that only
the required parts  of a cons cell are copied. The  behavior of the GC
algorithm is as described below,
\begin{algorithmic}
\STATE $new\_cell\gets$ dup\_heapcell(cell)
\IF{$celltype\ is\ cons\ and\ not(processed\_car\_part)$}
\STATE $new\_cdr\gets$ copy(cell.cdr)
\STATE $new\_cell.cdr\gets$ $new\_cdr$
\ENDIF
\end{algorithmic}
We   provide   an   intuition    about   the   correctness   of   this
algorithm. Consider the following facts about the print stack during a
GC,

>>>>

program shown, there are three such paths $ep_1 = \lbrace\pi_1, \pi_2, \pi_4 \rbrace$, 
$ep_2  =  \lbrace \pi_1, \pi_2,  \pi_5,  \pi_6,  \pi_7 \rbrace$,
 and $ep_3 = \lbrace \pi_3, \pi_2, \pi_1 \rbrace$.

The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
path to a common liveness environment, and a second map $\cal{P}$ maps
a program  point to the set of  e-paths on which  the point lies.
If a  program  point happens to fall  on more than one  such path, then
the  liveness environment of  the program  point is  the variable-wise
union of  the liveness environments  of the individual paths.   In the
example program, ${\cal M}(ep_1)= \lbrace \rbrace$, ${\cal M}(ep_2)= \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace  $ and
${\cal M}(ep_3)= \lbrace \px.\epsilon, \xl.\epsilon \rbrace $.
Since  ${\cal P}(\pi_1)$ is  $\lbrace ep_1,ep_2, ep_3 \rbrace$, 
the liveness  of $\pi_1$ is $\lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon,  \px.\epsilon  \rbrace$.  In summary,
our propagation of liveness is flow-insensitive but path-sensitive.
In the sequel
we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
$\cal{P}$ is easy and we do not elaborate it any further.

@@

\section{Generating Liveness Equations}
\begin{figure*}[t!]
%%\begin{picture}(100,130)(0,-90)
\begin{picture}(100,130)(0, -50)
  \begin{tabular}{cc}
    \begin{boxedminipage}{.51\textwidth}
      {\sf
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF$_{ep_3}$\ $\pi_3\!\!:\,$ \px
            \UNL{4} \hspace*{1.15cm} $\pi_4\!\!:\, (\RETURN_{ep_1}~0)$
            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\, (\RETURN_{ep_2}~(1~+~\py))$)))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_8\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
	  \UNL{2}   \hspace*{.78cm}   $\pi_9\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{3}   \hspace*{.78cm}    $\pi_{10}\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN~$\pi_{11}\!\!:\,$ (\RETURN$_{ep_4}$\ \pw)))))))
\end{uprogram}}
        } 
    \end{boxedminipage}
    & 
    \begin{boxedminipage}{.49\textwidth}
      {\sf
        \small
        \begin{tabular}{lc}
        $\cal{M_{\mbox{\tiny length}}} = $ 
            \begin{tabular}{lc}            
              $[$
              $ep_2 \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
              \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
              $ep_1 \mapsto \lbrace \rbrace , $\\
              $ep_3 \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
              $]$
            \end{tabular} \\
            $\cal{P_{\mbox{\tiny length}}} = [$
              \begin{tabular}{lc}
                $\pi_{\mbox{1}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
                $ \pi_{\mbox{2}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \\
                $ \pi_{\mbox{3}} \mapsto \lbrace ep_3 \rbrace, $ \
                $ \pi_{\mbox{4}} \mapsto \lbrace ep_1 \rbrace, $ \
                $ \pi_{\mbox{5}} \mapsto \lbrace ep_2 \rbrace, $ \\
                $ \pi_{\mbox{6}} \mapsto \lbrace ep_2 \rbrace, $ \
                $ \pi_{\mbox{7}} \mapsto \lbrace ep_2 \rbrace $ \\
                $]$
              \end{tabular}\\
              $\cal{M_{\mbox{\tiny main}}} = $ 
              \begin{tabular}{lc}            
                $[$
                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
                  $]$
              \end{tabular} \\
              $\cal{P_{\mbox{\tiny main}}} = [$
                \begin{tabular}{lc}
                  $\pi_{\mbox{8}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_{\mbox{9}} \mapsto \lbrace ep_4 \rbrace, $ \\
                  $ \pi_{\mbox{10}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_{\mbox{11}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \ldots  $ \
                  $]$  
                \end{tabular}
        \end{tabular}
      }
    \end{boxedminipage}
    \\
    (a) Example program.&
    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
        (b) {\em e-path} to liveness environment map \\
        (c) Prog pt to {\em e-paths } map
    \end{tabular}}
  \end{tabular}
\end{picture}
\kern -3ex
\vspace{0.5in}
\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
\end{figure*}

@@

\subsection{Generating Liveness Equations for Function bodies}
For the running example, $\length$ has calls from $\main$ at $\pi_{11}$
and a recursive call at $\pi_6$.
\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
So $\sigma_{\length} =
     \deltacall{\length}{\pi_{11}}{\main}  \cup \deltacall{\length}{\pi_6}{\length}$.
Filling in  the values gives:

@@

For the running example, containing unique GC points (corresponding to unique {\em e-paths} sets) $ep_1, ep_2, ep_3$ in $\length$ and
$ep_4$ in $e_\main$, this gives

\begin{align*}
\Lanv{7}{\xl} &= \acdr\Lf{\length}{1}{\bX\sigma_{\mathit{\length}}} \cup \bX\sigma_{\mathit{\length}} \\
\Lanv{7}{\pu} &= \Lf{\length}{1}{\bX\sigma_{\mathit{\length}}} \\
\Lanv{7}{\py} &= \bX\sigma_{\mathit{\length}} \\
\Lanv{3}{\xl} &= \bX\sigma_{\mathit{\length}} \\
\Lanv{3}{\px} &= \bX\sigma_{\mathit{\length}}
\\[1ex]
  \Lanv{9}{\pb}  &= \Lf{\length}{1}{\sigma_{\mathit{\!all}}} \\
  \Lanv{9}{\pw}  &= \sigma_{\mathit{\!all}}
\end{align*}

@@

\section{Implementation}
% GC assumptions
By the  time the program  starts executing, the liveness  automata are
created and stored in a file.   This is re-used every time the program
executes.
% GC points - mention reduction of gc points due to e-paths
An important  difference in GC of  lazy language compared  to an eager
language  is  the  point  of  garbage collection.  In  case  of  eager
languages only \CONS~ points and  function calls are the GC points. In
a lazy language  every \LET~ statement is a  potential GC point. Hence
we  need  to store  liveness  automata  corresponding  to every  \LET~
statement in the program. This leads  to an explosion in the number of
automata. From our observation that  along an {\em E-path} all program
points have  the same liveness, we  can reduce the  number of automata
state required  by sharing  the same automata  for all  program points
along an {\em E-path}. In the  example program, we only need to create
automata for program  points \ldots, since they are  the only distinct
{\em E-path}s in the program.
% Abstract description of the GC algorithm
%% \begin{algorithmic}
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \IF{$celltype\ is\ cons\ and\ not(copied\_using\_rgc)$}
%% \STATE $new\_car\gets$ copy($cell.car$)
%% \STATE $new\_cell.car\gets$ $new\_car$ 
%% \STATE $new\_cdr\gets$ copy(cell.cdr)
%% \STATE $new\_cell.cdr\gets$ $new\_cdr$
%% \ELSE
%% \IF {$celltype\ is\ closure\ and\ cell\ not\ already\ copied$}
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \STATE $new\_cell.arg1\gets$ copy($cell.arg1$)  
%% \STATE $new\_cell.arg2\gets$ copy($cell.arg2$)  
%% \ELSE
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \ENDIF
%% \ENDIF
%% \end{algorithmic}

@@

Another  difference that is  present in  GC of  lazy language  is the
presence  of  a print  stack.  In  lazy  languages the  evaluation  of
closures is  driven by the print  function.  Hence, during  a GC there
will be heap  references that will be present on  the stack which also
have  to  copied/collected.   One  solution  is  to  assume  that  all
references  present  on  the  print  stack are  fully  live  and  copy
everything pointed  by them  akin to reachability.   This is  safe but
copies heap  cells which  have already been  printed. Since  the print
function is  not part of  the executing program its  liveness automata
are not  considered. Therefore we  simulate the liveness of  the print
function by using a flag in  the heap cell. The flag ensures that only
the required parts  of a cons cell are copied. The  behavior of the GC
algorithm is as described below,
%% \begin{algorithmic}
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \IF{$celltype\ is\ cons\ and\ not(processed\_car\_part)$}
%% \STATE $new\_cdr\gets$ copy(cell.cdr)
%% \STATE $new\_cell.cdr\gets$ $new\_cdr$
%% \ENDIF
%% \end{algorithmic}
We   provide   an   intuition    about   the   correctness   of   this
algorithm. Consider the following facts about the print stack during a
GC,
